---
title: "Getting started with esmc"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with esmc}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
library(esmc)
library(ggplot2)
```

## Introduction

When comparing soil organic carbon (SOC) stocks — or any other soil property
whose stock varies with depth, such as total nitrogen, phosphorus, or other
elements — between treatments or over time, the standard approach is to sample
soil at fixed depths (e.g., 0–10, 10–30, 30–50 cm) and compare stocks within
those intervals. However, the mass of soil contained within a fixed-depth
interval depends on bulk density (BD). When BD differs between treatments —
due to compaction, tillage, organic matter accumulation, or other processes —
samples collected at the same depth contain different amounts of soil mass.
Comparing stocks across unequal soil masses can produce statistical artifacts:
apparent differences where none exist (Type I error inflation), or biased
estimates of true differences.

The **Equivalent Soil Mass (ESM) framework** corrects for this by standardizing
stocks to a common mineral soil mass reference rather than a common depth. For
each soil profile, a continuous curve of cumulative stock as a function of
cumulative mineral soil mass — C(M) — is fitted using interpolation, and stocks
are evaluated at a common reference mass across all profiles.

`esmc` provides a unified implementation of the ESM framework for any soil
property whose stock varies with depth. While SOC is used as the example
throughout this vignette, the same workflow applies to total nitrogen, 
phosphorus, or any other element or property of interest. The package includes
multiple interpolation methods, flexible reference mass options, diagnostic
tools, and an interactive simulator. It is based on the recommendations of
Villarino et al. (2026).

---

## Example dataset

Throughout this vignette we use a simulated dataset representing a paired
experiment with:

- 2 sites (`Site`: `"A"`, `"B"`)
- 2 years (`yr`: `2010`, `2020`)
- 2 treatments (`trt`: `"ctrl"` control, `"trt"` treatment with increased BD)
- 3 replicates per treatment (`rep`: 1, 2, 3)
- 5 depth intervals: 0–10, 10–30, 30–50, 50–70, 70–100 cm

The treatment simulates a 15% increase in bulk density near the surface
(logistic decay with depth), which is a realistic scenario in compaction or
tillage studies.

```{r simulate-data}
# Baseline BD profile (Wiese et al. function)
bd_wiese <- function(z_cm) {
  z_m <- (z_cm + 0.5) / 100
  0.1140 * log(z_m) + 1.23
}

# Generate one soil profile
make_profile <- function(site, yr, trt, rep, bd_inc_pct = 0, seed = NULL) {
  if (!is.null(seed)) set.seed(seed)
  layers <- list(c(0,10), c(10,30), c(30,50), c(50,70), c(70,100))

  core_mult <- exp(rnorm(1, mean = -0.5 * 0.07^2, sd = 0.07))

  do.call(rbind, lapply(layers, function(z) {
    z_mid   <- mean(z)
    bd_base <- bd_wiese(z_mid) * core_mult

    if (trt == "trt" && bd_inc_pct > 0) {
      f       <- plogis((50 - z_mid) / 6)
      bd_base <- bd_base * (1 + (bd_inc_pct / 100) * f)
    }

    bd_obs  <- pmax(bd_base + rnorm(1, 0, 0.03), 0.6)
    soc_true <- 2.5 * ((z_mid + 3) / 3)^(log(0.6/2.5) / log(103/3))
    soc_obs  <- soc_true * exp(rnorm(1, -0.5 * 0.05^2, 0.05))

    data.frame(Site  = site, yr = yr, trt = trt, rep = rep,
               upper = z[1], lower = z[2],
               bd    = round(bd_obs, 3),
               soc   = round(soc_obs, 3))
  }))
}

# Generate full dataset
set.seed(42)
combos <- expand.grid(
  site = c("A","B"), yr = c(2010, 2020),
  trt  = c("ctrl","trt"), rep = 1:3,
  stringsAsFactors = FALSE
)

df <- do.call(rbind, lapply(seq_len(nrow(combos)), function(i) {
  make_profile(
    site       = combos$site[i],
    yr         = combos$yr[i],
    trt        = combos$trt[i],
    rep        = combos$rep[i],
    bd_inc_pct = if (combos$trt[i] == "trt") 15 else 0,
    seed       = i * 17 + 3
  )
}))
rownames(df) <- NULL
```

The dataset has `r nrow(df)` rows (`r nrow(df)/5` profiles × 5 depth
intervals each):

```{r show-data}
head(df, 10)
```

### Required columns

`esm_compare()` expects a data frame with at minimum these columns:

| Column | Description | Units |
|--------|-------------|-------|
| `upper` | Upper boundary of depth interval | cm |
| `lower` | Lower boundary of depth interval | cm |
| `bd` | Bulk density | g cm⁻³ |
| `trt` | Treatment identifier | — |
| `soc` | Soil property to analyze | see `soil_prop_unit` |
| + grouping columns | e.g. `Site`, `yr`, `rep` | — |

---

## Basic usage

The main function is `esm_compare()`. At minimum it requires the data frame,
the name of the soil property column (`soil_prop`), and the columns that
uniquely identify each profile (`profile_by`):

```{r basic-usage}
result <- esm_compare(
  data           = df,
  soil_prop      = "soc",
  profile_by     = c("Site", "yr", "trt", "rep"),
  reference_mode = "min",
  reference_by   = c("Site", "yr"),
  interp         = "spline",
  soil_prop_unit = "pct",
  som_from_soc   = TRUE
)
```

The function returns a list. A summary of its elements:

```{r names}
names(result)
```

---

## Understanding the output

### `$layer` — ESM stock per depth layer

The most commonly used output. Each row is one depth layer for one profile,
with the SOC stock adjusted to the reference mineral soil mass.

```{r layer}
head(result$layer, 10)
```

**Columns:**

| Column | Description |
|--------|-------------|
| `Site`, `yr`, `trt`, `rep` | Profile identifiers (from `profile_by`) |
| `upper`, `lower` | Depth interval boundaries (cm) |
| `target_mass_upper_kg_m2` | Reference mineral mass at the upper boundary (kg m⁻²) |
| `target_mass_lower_kg_m2` | Reference mineral mass at the lower boundary (kg m⁻²) |
| `cum_upper_kg_m2` | Cumulative SOC stock at the upper boundary (kg C m⁻²) |
| `cum_lower_kg_m2` | Cumulative SOC stock at the lower boundary (kg C m⁻²) |
| `layer_stock_kg_m2` | ESM-adjusted SOC stock for this layer = `cum_lower - cum_upper` (kg C m⁻²) |
| `extrapolated` | `TRUE` if the target mass exceeded the observed mass range |

### `$targets` — Reference mass at each depth boundary

The reference mineral soil mass used at each depth boundary. This is the
common mass to which all profiles are standardized.

```{r targets}
result$targets
```

**Columns:**

| Column | Description |
|--------|-------------|
| `Site`, `yr` | Stratum identifiers (from `reference_by`) |
| `lower` | Depth boundary (cm) |
| `target_mass_kg_m2` | Reference mineral soil mass at this boundary (kg m⁻²) |

With `reference_mode = "min"`, the target at each boundary is the minimum
cumulative mineral mass observed across all profiles within the stratum —
this guarantees interpolation (no extrapolation) for all profiles.

### `$cumulative` — Cumulative ESM stock at the reference mass

The total SOC stock from the surface down to the reference mass, for each
profile.

```{r cumulative}
head(result$cumulative)
```

**Columns:**

| Column | Description |
|--------|-------------|
| `Site`, `yr`, `trt`, `rep` | Profile identifiers |
| `target_mass_kg_m2` | Reference mineral soil mass (kg m⁻²) |
| `cum_stock_kg_m2` | Cumulative SOC stock at the reference mass (kg C m⁻²) |
| `extrapolated` | `TRUE` if extrapolation was required |

### `$diagnostics` — Per-profile interpolation diagnostics

Useful for checking whether any profiles required extrapolation and why.

```{r diagnostics}
head(result$diagnostics)
```

**Columns:**

| Column | Description |
|--------|-------------|
| `Site`, `yr`, `trt`, `rep` | Profile identifiers |
| `interp` | Interpolation method used |
| `target_mass_kg_m2` | Reference mineral soil mass (kg m⁻²) |
| `max_cum_mass_kg_m2` | Maximum cumulative mineral mass observed for this profile (kg m⁻²) |
| `extrapolated` | `TRUE` if `target_mass > max_cum_mass` |
| `note` | Warning note (e.g. `INSUFFICIENT_POINTS_SPLINE`) or `NA` |

### `$profile_mass` — C(M) curve nodes per profile

The data points used to fit the C(M) interpolation curve for each profile:
cumulative mineral soil mass vs cumulative SOC stock, including the (0, 0)
anchor point. Primarily used for diagnostic plotting.

```{r profile-mass}
head(result$profile_mass, 8)
```

**Columns:**

| Column | Description |
|--------|-------------|
| `Site`, `yr`, `trt`, `rep` | Profile identifiers |
| `cum_mineral_mass_kg_m2` | Cumulative mineral soil mass (kg m⁻²) |
| `cum_stock_kg_m2` | Cumulative SOC stock (kg C m⁻²) |

### `$interval` — Per-interval computations

Intermediate computations for each depth interval: soil mass, mineral mass,
and SOC stock before ESM adjustment.

```{r interval}
head(result$interval)
```

**Columns:**

| Column | Description |
|--------|-------------|
| `Site`, `yr`, `trt`, `rep` | Profile identifiers |
| `upper`, `lower` | Depth interval boundaries (cm) |
| `bd` | Bulk density (g cm⁻³) |
| `soil_mass_kg_m2` | Total soil mass in interval (kg m⁻²) |
| `mineral_mass_kg_m2` | Mineral soil mass in interval (kg m⁻²) |
| `cum_mineral_mass_kg_m2` | Cumulative mineral soil mass at bottom of interval (kg m⁻²) |
| `stock_kg_m2` | SOC stock in interval (kg C m⁻²) |
| `cum_stock_kg_m2` | Cumulative SOC stock at bottom of interval (kg C m⁻²) |

### `$meta` — Call metadata

A list recording all settings used in the call, useful for reproducibility.

```{r meta}
str(result$meta)
```

---

## Diagnostic plot

`esm_plot_spline()` visualizes the C(M) curve for each profile: observed
nodes as points and the fitted interpolation curve as a line. A vertical
dashed line marks the reference mass. 

**Note:** `esm_plot_spline()` requires `$diagnostics` to be present in the
output of `esm_compare()`. This is included when `output = "both"` (default)
or `output = "cumulative"`, but **not** when `output = "layers"`.

- **Blue line** — target is within the observed data range (interpolation)
- **Red line + shading** — target exceeds the observed data range (extrapolation)
- **Dashed red curve** — the extrapolated portion of the C(M) curve

```{r plot-spline, fig.width = 9, fig.height = 7}
esm_plot_spline(result, profile_by = c("Site", "yr", "trt", "rep"))
```

This plot is particularly useful for understanding why ESM estimates can be
unreliable when the reference mass exceeds the maximum observed mass of a
profile — the spline must extrapolate beyond the data, which can introduce bias.

---

## Reference modes

The `reference_mode` argument controls how the reference mineral soil mass
is defined. This is one of the most important decisions in ESM analysis.

### `"min"` — minimum observed mass (recommended)

The reference mass at each depth boundary is set to the **minimum** cumulative
mineral mass observed across all profiles within each stratum (`reference_by`).
This guarantees that all profiles can be evaluated by interpolation — no
extrapolation is ever needed.

```{r ref-min}
res_min <- esm_compare(
  data           = df,
  soil_prop      = "soc",
  profile_by     = c("Site", "yr", "trt", "rep"),
  reference_mode = "min",
  reference_by   = c("Site", "yr"),
  interp         = "spline",
  soil_prop_unit = "pct",
  som_from_soc   = TRUE
)
# No extrapolation
sum(res_min$layer$extrapolated)
```

### `"ref_trt"` — reference treatment mass

The reference mass is set to the **mean** cumulative mineral mass of a
designated reference treatment. Profiles from other treatments may require
extrapolation if their mass is lower than the reference treatment's mass.

```{r ref-trt}
res_reftrt <- esm_compare(
  data           = df,
  soil_prop      = "soc",
  profile_by     = c("Site", "yr", "trt", "rep"),
  reference_mode = "ref_trt",
  ref_trt        = "ctrl",
  reference_by   = c("Site", "yr"),
  interp         = "spline",
  soil_prop_unit = "pct",
  som_from_soc   = TRUE
)
# Some extrapolation may occur in treatment profiles
sum(res_reftrt$layer$extrapolated)
```

### `"custom"` — user-supplied mass

The reference mass is set to a fixed value supplied by the user. Useful when
comparing results across studies or when a specific mass is theoretically
motivated.

```{r ref-custom}
res_custom <- esm_compare(
  data           = df,
  soil_prop      = "soc",
  profile_by     = c("Site", "yr", "trt", "rep"),
  reference_mode = "custom",
  custom_mass    = 900,
  interp         = "spline",
  soil_prop_unit = "pct",
  som_from_soc   = TRUE
)
res_custom$targets
```

---

## Interpolation methods

The `interp` argument controls how the C(M) curve is fitted between observed
nodes. The choice matters most when the reference mass falls between two
observed points (interpolation) or beyond the last observed point
(extrapolation).

### `"spline"` — cubic spline (recommended)

Fits a natural cubic spline through the C(M) nodes. Captures the curvature
in the SOC profile, which is common in most soils — SOC decreases non-linearly
with depth and mass. **Recommended when ≥ 3 depth intervals are available.**

```{r interp-spline}
res_spline <- esm_compare(
  data = df, soil_prop = "soc",
  profile_by     = c("Site", "yr", "trt", "rep"),
  reference_mode = "min", reference_by = c("Site", "yr"),
  interp = "spline", soil_prop_unit = "pct", som_from_soc = TRUE
)
```

### `"linear"` — linear interpolation

Connects C(M) nodes with straight line segments. Simpler and more conservative
than spline, but may underestimate curvature. **Recommended when only 1 depth
interval is available.** With 2 or more intervals, the (0, 0) anchor counts as
an additional point, giving the spline sufficient points to fit. In practice,
spline generally reduces systematic errors further than linear when multiple
layers are available, though performance depends on the depthwise SOC
distribution of the soil (Potash et al., 2026).

```{r interp-linear}
res_linear <- esm_compare(
  data = df, soil_prop = "soc",
  profile_by     = c("Site", "yr", "trt", "rep"),
  reference_mode = "min", reference_by = c("Site", "yr"),
  interp = "linear", soil_prop_unit = "pct", som_from_soc = TRUE
)
```

### Comparing methods

```{r compare-methods, fig.width = 7, fig.height = 4}
# Extract layers for Site A, yr 2010, layers 0-10 and 10-30
get_layer <- function(res, method_name) {
  sub <- res$layer[res$layer$Site  == "A" &
                   res$layer$yr    == 2010 &
                   res$layer$lower %in% c(10, 30), ]
  sub$method <- method_name
  sub
}

df_comp <- rbind(
  get_layer(res_spline, "spline"),
  get_layer(res_linear, "linear")
)

agg <- aggregate(layer_stock_kg_m2 ~ method + trt + upper + lower,
                 data = df_comp, FUN = mean)
agg$layer <- paste0(agg$upper, "-", agg$lower, " cm")

ggplot(agg, aes(x = method, y = layer_stock_kg_m2, fill = trt)) +
  geom_col(position = "dodge") +
  facet_wrap(~ layer) +
  labs(x     = "Interpolation method",
       y     = expression("SOC stock (kg C m"^{-2}*")"),
       fill  = "Treatment",
       title = "ESM stocks by interpolation method (Site A, 2010)") +
  theme_bw()
```

---

## The Shiny app

`run_esmc_app()` launches an interactive simulator for exploring how bulk
density changes affect statistical performance when comparing SOC stocks
using fixed-depth (FD) vs ESM approaches.

```{r run-app, eval = FALSE}
run_esmc_app()
```

The app allows you to:

- Switch between **H0** (no true SOC change — Type I error focus) and
  **H1** (true SOC change — power focus) scenarios
- Control BD change magnitude, SOC stratification profile, and measurement
  noise
- Run Monte Carlo simulations to estimate Type I error rates, bias, and
  statistical power
- Generate power curves as a function of sample size (number of cores)

This is useful for study design — understanding how many cores are needed
to detect a given SOC change with adequate power using ESM vs fixed-depth
methods.

> **Note:** `run_esmc_app()` requires `shiny`, `ggplot2`, `future`, and
> `future.apply`. Install them with:
> ```r
> install.packages(c("shiny", "ggplot2", "future", "future.apply"))
> ```

---

## Reference

Villarino, S.H., Castellano, M., and Miguez, F.E. (2026). Towards a Unified
Equivalent Soil Mass Framework. *In review*.

Potash, E., von Haden, A.C., Bergschneider, L., Douglass, M.S., Ferrari, L.,
Ibarra Lopez, R., Reconco Martinez, A., and Margenot, A.J. (2026). Empirical
evaluation and recommendations for equivalent soil mass adjustment of soil
organic carbon stocks. *Geoderma*, 469, 117801.
https://doi.org/10.1016/j.geoderma.2026.117801
